\documentclass[12pt, a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{titlesec}
\usepackage{xcolor}
\usepackage{parskip}

% Page Geometry
\geometry{
 a4paper,
 total={170mm,257mm},
 left=25mm,
 top=25mm,
 bottom=25mm,
}

% Styling
\definecolor{familyblue}{RGB}{0, 80, 150}
\titleformat{\section}{\large\bfseries\color{familyblue}}{}{0em}{}
\titleformat{\subsection}{\bfseries}{}{0em}{}

\title{\textbf{The Team Protocol (Family Edition v2.0)}\\ \large Our "User Manual" for working together}
\author{Based on the General Theory of Collaboration}
\date{\today}

\begin{document}

\maketitle

\begin{center}
    \textit{These are not rules to punish you. These are agreements to help us understand each other without the noise.}
\end{center}

\hrule
\vspace{1em}

\section{Phase 1: The Check-In (Are we aligned?)}
\textit{(Theory: Vector Alignment \& Constraints)}

Before we start something big (or if things feel weird), we check two things:

\subsection{1. The Compass Check (The Goal)}
We cannot work together if you are pulling left and I am pulling right.
\begin{itemize}
    \item \textbf{Question:} "What Mode are we in?"
    \item \textbf{The Modes:}
    \begin{itemize}
        \item \textbf{Fun Mode:} We are being silly and playing.
        \item \textbf{Focus Mode:} We need to get things done (homework/catching a train).
        \item \textbf{Rest Mode:} Low energy, quiet time.
    \end{itemize}
    \item \emph{Rule: If we are in different modes, we stop and fix that first.}
\end{itemize}

\subsection{2. The Weather Report (The Pre-Mortem)}
Is there a storm coming? Tell us \emph{before} it rains.
\begin{itemize}
    \item \textbf{Question:} "How is your Battery?" (0\% to 100\%)
    \item \textbf{Question:} "Do you feel a storm coming?"
    \item \emph{Answer:} "Yes, I'm tired, so I have a short fuse. It's not your fault, but please be gentle."
\end{itemize}

\section{Phase 2: The Signals (How we talk)}
\textit{(Theory: Impedance Matching \& Querying)}

Sometimes we misunderstand each other. We use labels to fix this.

\subsection{1. The Label}
Don't let me guess what you mean. Tell me what the message is.
\begin{itemize}
    \item \textbf{"Just Venting":} I am complaining to get it out. I don't need you to fix it. Just listen.
    \item \textbf{"Fact Check":} I am telling you something true (like a time or a rule).
    \item \textbf{"Feeling":} I am telling you how I feel (it might not be logical, but it's real).
\end{itemize}

\subsection{2. The Mirror (Checking Blind Spots)}
Sometimes I am grumpy and I don't know it.
\begin{itemize}
    \item \textbf{The Signal:} You can say: \textbf{"Check your Mirror."}
    \item \textbf{The Rule:} I promise not to get mad. I will stop for 5 seconds and ask myself: \emph{"Do I sound grumpy?"}
    \item \emph{Note: The mirror is always right. We don't get mad at the mirror.}
\end{itemize}

\section{Phase 3: Safety Controls (Stopping the crash)}
\textit{(Theory: Latency \& State Synchronization)}

When things go wrong, we use these buttons.

\subsection{1. The "Echo" (Immediate Ack)}
If I say "Stop" or "Careful," I need to know you heard me.
\begin{itemize}
    \item \textbf{Rule:} You don't just stop; you look at me and nod (or say "Okay").
    \item \textbf{Why:} If you don't Echo, I think you didn't hear me, and I will get louder.
\end{itemize}

\subsection{2. The "Save Point" (Syncing Up)}
Sometimes we remember things differently.
\begin{itemize}
    \item \textbf{Action:} If we are confused, we press "Save."
    \item \textbf{Say:} "Let's save the game. We agree that [X] happened, right?"
    \item \textbf{Why:} So we don't fight about what happened 10 minutes ago.
\end{itemize}

\section{Phase 4: The Update (Learning)}
\textit{(Theory: Plasticity)}

If we keep fighting about the same thing, trying harder won't help.
\begin{itemize}
    \item \textbf{The System Update:} We say: "This rule isn't working. Let's make a new one."
    \item \emph{Example:} If we always fight about shoes, we change where the shoes live. We don't just yell about shoes.
\end{itemize}

\vspace{2em}
\hrule
\begin{center}
    \textit{Signed by the Team:}
\end{center}

\end{document}