\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\title{Introducing the Discovarians: Naming the Learning Networks}
\author{Albert Jan van Hoek}
\date{\today}

\begin{document}

\maketitle

When you discover something, you get the pleasure of naming it.

I chose the name for the \textbf{Theory of Long-term Collaboration} carefully. Its acronym, TLC, already exists—Tender Loving Care—which, in a way, is the logical conclusion of TLC itself. So the name seemed perfect.

\section*{The Identity of Learning Networks}

However, within TLC, the participants are learning networks. To apply TLC in real life, you first have to identify as one. And this identity needed a name.

In an earlier post, I wrote: \textit{I learn, therefore I am.}
After \textit{I think, therefore I am}—or \textit{cogito ergo sum}—something we all know.
To my great joy, \textit{I learn, therefore I am} in Latin is \textit{disco ergo sum}. The same ``disco'' as in discovery.

Hence, someone who identifies as a learning network is a \textbf{Discovarian}.

\section*{Spaces for Learning and Collaboration}

Discovarians meet in \textbf{Discothecs}—``disco'' for learning, ``thec'' for place; like a bibliothec. A space for learning, discovery, and collaboration.

\section*{Finding the Groove}

Within TLC, the goal is to find a routine between two learning systems that can sustain collaboration. This is an eternal chain, an eternal loop. In disco music, finding a loop that’s enjoyable—and keeps you dancing—is called a groove.

So, in TLC, finding and keeping that routine is called \textbf{``finding the groove.''} If you help others maintain that groove, you’re a \textbf{groovekeeper}.

\section*{A Disco Song and a Preaching}

Fittingly, let me share something I created last year: a disco song called \textit{Large-language Mindset}, inspired by treating each other like large-language models—totally neutral: no gender, no nationality, no other excuse—just learning.

Given the world today, as your newly appointed groovekeeper, my preaching is simple:
\begin{itemize}
    \item Don’t be anxious—believe in learning, and remember:
    \item The only place we’re allowed to drop body-moving shizzle is on the dance floor!
\end{itemize}

\end{document}
