%% ============================================================
%%  Artificial Intelligence and Public Health
%%  Rethinking Infectious Disease Surveillance
%%  Albert Jan van Hoek
%%  Version 3
%% ============================================================

\documentclass[aspectratio=169, 10pt]{beamer}

\usetheme{Madrid}
\usecolortheme{default}

% ── Colours ──────────────────────────────────────────────────
\definecolor{primary}{HTML}{065A82}
\definecolor{accent}{HTML}{F4A261}
\definecolor{lightbg}{HTML}{F5F9FC}
\definecolor{darktext}{HTML}{1A1A2E}

\setbeamercolor{palette primary}{bg=primary, fg=white}
\setbeamercolor{palette secondary}{bg=primary!80, fg=white}
\setbeamercolor{palette tertiary}{bg=primary!60, fg=white}
\setbeamercolor{palette quaternary}{bg=primary, fg=white}
\setbeamercolor{structure}{fg=primary}
\setbeamercolor{background canvas}{bg=lightbg}
\setbeamercolor{normal text}{fg=darktext}
\setbeamercolor{frametitle}{bg=primary, fg=white}
\setbeamercolor{block title}{bg=primary, fg=white}
\setbeamercolor{block body}{bg=primary!8, fg=darktext}
\setbeamercolor{alerted text}{fg=accent}

\setbeamertemplate{navigation symbols}{}

% ── Packages ─────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning}

% ── Footer ───────────────────────────────────────────────────
\setbeamertemplate{footline}{%
  \leavevmode%
  \hbox{%
    \begin{beamercolorbox}[wd=.5\paperwidth, ht=2.5ex, dp=1.5ex,
      left, leftskip=1em]{palette primary}
      \usebeamerfont{author in head/foot}Albert Jan van Hoek — RIVM
    \end{beamercolorbox}%
    \begin{beamercolorbox}[wd=.4\paperwidth, ht=2.5ex, dp=1.5ex,
      center]{palette secondary}
      \usebeamerfont{title in head/foot}\insertshorttitle
    \end{beamercolorbox}%
    \begin{beamercolorbox}[wd=.1\paperwidth, ht=2.5ex, dp=1.5ex,
      right, rightskip=1em]{palette primary}
      \insertframenumber/\inserttotalframenumber
    \end{beamercolorbox}%
  }%
}

% ── Title ────────────────────────────────────────────────────
\title[AI \& Public Health]{Artificial Intelligence and Public Health}
\subtitle{Rethinking Infectious Disease Surveillance}
\author{Albert Jan van Hoek}
\institute{RIVM — Dutch National Institute for Public Health and the Environment}
\date{}

% =============================================================
\begin{document}

% ── Title slide ───────────────────────────────────────────────
\begin{frame}[plain]
  \vspace{1.5cm}
  \titlepage
\end{frame}

% ── Who am I ─────────────────────────────────────────────────
\begin{frame}{Who am I?}
  \begin{itemize}
    \item Senior researcher at \textbf{RIVM}
    \item Principal Investigator, \textbf{Infectieradar} —
          participatory surveillance, European Influenzanet network
    \item Health economist and infectious disease modeller
    \item Working on NLP, agentic AI, and transformer modelling
    \item Field experience in \textit{Gambia, Kenya, Sierra Leone}
    \item Background: \textbf{Economics / Biology / Philosophy}
  \end{itemize}
\end{frame}

% =============================================================
\section{The misconception}

\begin{frame}{How we think about AI}
  \begin{block}{}
    AI is a tool.\\[0.4em]
    We deploy it. We control it. It executes tasks.
  \end{block}

  \pause
  \bigskip

  This framing shapes how we build surveillance systems.\\
  How we staff teams. How we evaluate risk.\\[0.5em]

  \begin{alertblock}{}
    \centering
    What if that framing is wrong — not just incomplete, but wrong?
  \end{alertblock}
\end{frame}

\begin{frame}{AI systems change as they are used}
  \begin{itemize}
    \item They generate outputs.
    \item Those outputs are evaluated.
    \item The system is updated.
  \end{itemize}

  \bigskip

  That is not execution. That is adaptation.\\[0.5em]

  \begin{alertblock}{}
    \centering
    The tool framing misses something structural.\\
    Let's be precise about what.
  \end{alertblock}
\end{frame}

% =============================================================
\section{The structure}

\begin{frame}{A three-part loop}
  \begin{center}
  \begin{tikzpicture}[
    node distance=3cm,
    box/.style={rectangle, rounded corners=4pt, draw=primary, thick,
                fill=primary!10, text width=2.8cm, align=center,
                minimum height=1.1cm, font=\bfseries\small},
    arr/.style={-{Stealth[scale=1.2]}, thick, color=primary}
  ]
    \node[box] (G) {Generate\\variation};
    \node[box, right of=G] (T) {Test against\\reality};
    \node[box, right of=T] (U) {Update\\structure};
    \draw[arr] (G) -- (T);
    \draw[arr] (T) -- (U);
    \draw[arr] (U.south) -- ++(0,-0.7) -| (G.south);
  \end{tikzpicture}
  \end{center}

  \bigskip
  Three components. Specific relations between them. One feedback loop.\\[0.4em]
  \textit{This is a functional definition, not a description.}
\end{frame}

\begin{frame}{When are two processes the same — not just similar?}
  Analogy is cheap. Structural identity is not.\\[0.6em]

  Two processes are \textbf{structurally identical} when they share:
  \begin{enumerate}
    \item the same functional components
    \item in the same relations to each other
    \item producing the same type of output
  \end{enumerate}

  \bigskip
  \begin{alertblock}{}
    \centering
    This is a testable criterion.\\
    We will apply it to evolution and to AI.
  \end{alertblock}
\end{frame}

\begin{frame}{Evolution — does it pass?}
  \begin{columns}[T]
    \column{0.58\textwidth}
      \begin{tabular}{lll}
        \toprule
        Component & Required & In evolution \\
        \midrule
        Generate  & variation produced       & genetic / phenotypic \\
        Test      & confronted with reality  & selection pressure \\
        Update    & structure changes        & differential reproduction \\
        \midrule
        \multicolumn{2}{l}{Same relations?}  & \checkmark \\
        \multicolumn{2}{l}{Same output type?} & \checkmark \\
        \bottomrule
      \end{tabular}

    \column{0.38\textwidth}
      \begin{alertblock}{\checkmark\ Passes}
        Evolution instantiates\\
        all three components\\
        in the right relations.\\[0.3em]
        Same process.\\
        Different substrate.
      \end{alertblock}
  \end{columns}
\end{frame}

\begin{frame}{AI learning — does it pass?}
  \begin{columns}[T]
    \column{0.58\textwidth}
      \begin{tabular}{lll}
        \toprule
        Component & Required & In AI training \\
        \midrule
        Generate  & variation produced       & outputs / weight space \\
        Test      & confronted with reality  & loss on data \\
        Update    & structure changes        & gradient descent \\
        \midrule
        \multicolumn{2}{l}{Same relations?}  & \checkmark \\
        \multicolumn{2}{l}{Same output type?} & \checkmark \\
        \bottomrule
      \end{tabular}

    \column{0.38\textwidth}
      \begin{alertblock}{\checkmark\ Passes}
        AI learning instantiates\\
        all three components\\
        in the right relations.\\[0.3em]
        Same process.\\
        Different substrate.
      \end{alertblock}
  \end{columns}
\end{frame}

\begin{frame}{The obvious objection}
  \begin{block}{``At that level of abstraction, anything fits.
                  Rivers flow. Markets clear.''}
    Good objection. Let's apply the criterion strictly.
  \end{block}

  \bigskip
  \begin{columns}[T]
    \column{0.5\textwidth}
      \begin{tabular}{lcc}
        \toprule
        System        & Internal model? & Updates structure? \\
        \midrule
        River         & \texttimes      & \texttimes \\
        Thermostat    & \texttimes      & \texttimes \\
        Market price  & \texttimes      & \texttimes \\
        \midrule
        Immune system & \checkmark      & \checkmark \\
        Evolution     & \checkmark      & \checkmark \\
        AI learning   & \checkmark      & \checkmark \\
        \bottomrule
      \end{tabular}

    \column{0.46\textwidth}
      A river has no internal model.\\
      A thermostat has a fixed setpoint — it does not update.\\[0.5em]
      \begin{alertblock}{}
        The criterion is discriminating.\\
        Most dynamic systems\\
        do \textit{not} pass it.
      \end{alertblock}
  \end{columns}
\end{frame}

% =============================================================
\section{Recursion}

\begin{frame}{One step further}
  Some systems do more than adapt.\\[0.5em]
  They change \textit{how} they adapt.

  \bigskip
  \begin{block}{Recursion: the loop operating on itself}
    \begin{itemize}
      \item Evolution: mutation rates are themselves under selection
      \item Humans: we improve our methods, not just our answers
      \item AI: fine-tuning, RLHF, emerging agentic architectures
    \end{itemize}
  \end{block}

  \bigskip
  \begin{alertblock}{}
    \centering
    Learning improves learning.\\
    That is not a metaphor either.
  \end{alertblock}
\end{frame}

% =============================================================
\section{Substrate}

\begin{frame}{The loop needs something to run on}
  The substrate is not a container.\\
  It determines which components are possible.

  \bigskip
  \begin{tabular}{lll}
    \toprule
    Substrate            & Makes possible            & Loop running on \\
    \midrule
    Host population      & replication \& selection  & viral evolution \\
    Training data        & error signal \& update    & AI learning \\
    Interaction network  & variation \& feedback     & collective intelligence \\
    \bottomrule
  \end{tabular}

  \bigskip
  \begin{alertblock}{}
    \centering
    Change the substrate, and you change the loop.\\
    \textbf{You do not need to change the elements.}
  \end{alertblock}
\end{frame}

\begin{frame}{The real question}
  If the loop is real, and substrate-dependent, then the question changes.

  \bigskip
  \begin{columns}[c]
    \column{0.44\textwidth}
      \begin{block}{The wrong question}
        \centering
        How powerful is the AI?\\[0.3em]
        \small Asks about one component in isolation.
      \end{block}

    \column{0.52\textwidth}
      \begin{alertblock}{The right question}
        \centering\large
        \textbf{How well does your network learn?}\\[0.3em]
        \normalsize
        This is a precise, technical question.\\
        And it applies to your AI,\\
        your surveillance system,\\
        and your organisation equally.
      \end{alertblock}
  \end{columns}
\end{frame}

% =============================================================
\section{Public health}

\begin{frame}{An epidemic is already a learning system}
  \begin{itemize}
    \item Pathogens generate variation continuously.
    \item Populations and immune systems test those variants.
    \item Both sides update — pathogen and host.
  \end{itemize}

  \bigskip
  Transmission is not a property of the virus.\\
  It is a property of the \textit{network}.

  \bigskip
  \begin{alertblock}{}
    \centering
    We are already inside a loop.\\
    The question is whether our surveillance system is too.
  \end{alertblock}
\end{frame}

\begin{frame}{Surveillance today}
  \begin{block}{}
    \centering
    Data $\;\rightarrow\;$ Analysis $\;\rightarrow\;$ Decision
  \end{block}

  \bigskip
  A pipeline. Linear. Slow.\\[0.5em]

  The loop is broken at two points:
  \begin{itemize}
    \item no feedback from decision back to observation
    \item no structural update — the system does not learn from its own output
  \end{itemize}
\end{frame}

\begin{frame}{What adaptive surveillance looks like}
  \begin{alertblock}{}
    \centering\large
    Observe $\;\rightarrow\;$ Interpret $\;\rightarrow\;$ Act
    $\;\rightarrow\;$ Update $\;\rightarrow\;$ Repeat
  \end{alertblock}

  \bigskip
  The difference is not speed. It is structure.\\[0.5em]
  A surveillance system that learns from what it detects\\
  is not a faster pipeline — it is a different kind of system.
\end{frame}

% =============================================================
\section{AI as amplifier}

\begin{frame}{What AI adds to the loop}
  \begin{itemize}
    \item faster generation of hypotheses
    \item large-scale testing against data
    \item rapid structural updating
  \end{itemize}

  \bigskip
  \begin{alertblock}{}
    \centering
    AI does not replace the loop.\\
    It runs the loop faster.
  \end{alertblock}
\end{frame}

\begin{frame}{AI failures — read through the same lens}
  \begin{tabular}{lll}
    \toprule
    Failure mode   & Loop component & What breaks \\
    \midrule
    Hallucination  & Test           & Reality check is weak or absent \\
    Bias           & Generate       & Input variation is distorted \\
    Drift          & Update         & Structural update becomes unstable \\
    \bottomrule
  \end{tabular}

  \bigskip
  \begin{block}{}
    The framework does not just explain what AI does well.\\
    It localises exactly where it fails — and why.
  \end{block}
\end{frame}

% =============================================================
\section{Building the system}

\begin{frame}{What I am building at RIVM}
  \begin{columns}[T]
    \column{0.48\textwidth}
      \begin{block}{Faster generation}
        \begin{itemize}
          \item NLP for questionnaire analysis
          \item Synthetic populations
          \item Transformer modelling
        \end{itemize}
      \end{block}
      \begin{block}{Better testing}
        \begin{itemize}
          \item Benchmarking AI outputs
          \item Agentic quality control
        \end{itemize}
      \end{block}

    \column{0.48\textwidth}
      \begin{block}{Structural update}
        \begin{itemize}
          \item FAIR and open data
          \item Digital longitudinal data
          \item Open source infrastructure
        \end{itemize}
      \end{block}
      \begin{alertblock}{}
        Each project targets\\
        one component of the loop.
      \end{alertblock}
  \end{columns}
\end{frame}

\begin{frame}{Pandemic preparedness is not a project}
  It is a continuous mission.

  \bigskip
  The right frame: a loop that improves itself,\\
  with internal regulation of that improvement.

  \bigskip
  \begin{itemize}
    \item logistics, IT, protocols, privacy, communication, training
  \end{itemize}

  \bigskip
  \begin{alertblock}{}
    \centering
    \textit{k}-cover: the network is covered from every direction of failure.\\
    Not a state to reach. A property to maintain.
  \end{alertblock}
\end{frame}

\begin{frame}{Maintaining the loop at the human level}
  Feedback loops break when people stop giving honest signal.\\[0.5em]
  That is not a values claim. It is a systems claim.

  \bigskip
  \begin{columns}[T]
    \column{0.46\textwidth}
      \begin{block}{Structurally}
        More interaction.\\
        More feedback.\\
        More collaboration.
      \end{block}

    \column{0.50\textwidth}
      \begin{alertblock}{Functionally required}
        Honesty. Openness. Repair.\\[0.4em]
        \small Not virtues bolted onto the system.\\
        The conditions under which the\\
        feedback loop stays intact.
      \end{alertblock}
  \end{columns}
\end{frame}

% =============================================================
\section{Conclusion}

\begin{frame}{What you can do}
  \begin{enumerate}
    \item \textbf{Use AI — seriously.}
          It extends the loop you already run.
    \bigskip
    \item \textbf{Think at the network level.}
          Your node matters less than your connections.
    \bigskip
    \item \textbf{Protect your domain knowledge.}
          Infectious disease expertise built over years\\
          is not replicable quickly — pair it with data literacy.
    \bigskip
    \item \textbf{Collaborate deliberately.}
          Collective capability is not automatic.\\
          It degrades without maintenance.
  \end{enumerate}
\end{frame}

\begin{frame}{Conclusion}
  \begin{itemize}
    \item The loop — generate, test, update — is a real structural criterion.
    \item Evolution, learning, and AI instantiate it. Rivers do not.
    \item The substrate determines what the loop can do.
    \item AI accelerates the loop. It does not replace it.
    \item Public health surveillance needs to \textit{become} the loop.
  \end{itemize}

  \bigskip
  \begin{alertblock}{}
    \centering
    We are not building tools.\\
    \textbf{We are building systems that learn.}\\[0.3em]
    The difference matters.
  \end{alertblock}
\end{frame}

% ── Closing slide ─────────────────────────────────────────────
\begin{frame}[plain]
  \begin{center}
    \vspace{1.2cm}
    {\large\textit{Viral wisdom:}}\\[0.8em]
    {\Large Keep alive what keeps you alive.}\\[0.4em]
    {\normalsize And improve it for the next generation.}\\[1.8em]
    \rule{0.55\textwidth}{0.4pt}\\[1em]
    {\LARGE\textbf{\textit{Disco ergo sum.}}}\\[0.3em]
    {\normalsize I learn, therefore I am.}
  \end{center}
\end{frame}

\end{document}