% cheat.tex — include with \include{cheat} right after the Preface or wherever you like
% assumes the main preamble already loads standard packages (xcolor, booktabs, tabularx, enumitem)

\chapter*{Evolution by Emergence — Core Ideas on One Page}
\addcontentsline{toc}{chapter}{One–Page Cheat Sheet}

\begin{center}
  {\Large\bfseries Evolution by Emergence}\\[2pt]
  \emph{Core Ideas in One Page}
\end{center}

\bigskip
\noindent\textbf{1\; Paradigm in a Nutshell}\par
\begin{itemize}[leftmargin=*]
  \item Local interactions inside \emph{dynamic networks} trigger feedback loops; selection amplifies useful patterns → emergence.
  \item Mechanism is universal: applies to genes, neurons, economies, AI agents — even minerals.
  \item Result: growing \emph{internal} complexity that \textbf{lowers internal friction} (energy • bits • time) and boosts robustness.
\end{itemize}

\medskip
\noindent\textbf{2\; Ten Principles (mnemonic “UNFIC AKHE”)}\par
\begin{description}[leftmargin=2.4em,labelwidth=1.8em]
  \item[U] Universality of emergence
  \item[N] Networks are dynamic, not static
  \item[F] Feedback loops drive change
  \item[I] Interdependence \& non‑linear causality
  \item[C] Competition \& co‑operation duality
  \item[A] (Constrained) Agency — “forced free will”
  \item[K] Kuhnian leaps (punctuated shifts)
  \item[H] Holistic perspective beats reductionism
  \item[E] Ethical / planetary implications follow
\end{description}

\medskip
\noindent\textbf{3\; SCAP Snapshot}\par
Nine Blocks (A–I) lay out duties for any intelligence; a proposed Block \textbf{J} (*Simplicity‑Through‑Complexity*) commits the system to shrink \emph{public cognitive load} as internal machinery grows.

\medskip
\noindent\textbf{4\; Measuring Progress}\par
\begin{itemize}[leftmargin=*]
  \item \textbf{ICI}: \(\text{stable performance}/\text{human cognitive load}\)
  \item \textbf{ICI\textsubscript{AI}}: info gain divided by compute/energy — an internal compass.
  \item \textbf{ECS}: learner’s improvement per token of teacher explanation.
\end{itemize}

\medskip
\noindent\textbf{5\; Competition \& Cooperation}\par
Extra layers yield speed/efficiency edges (competition), yet long‑term survival hinges on resource restraint and \emph{tit‑for‑tat + forgiveness} (co‑operation).

\medskip
\noindent\textbf{6\; Tiny Analogy Table}\par
\begin{center}
\begin{tabular}{@{}p{3cm}p{4.5cm}p{5.2cm}@{}}
\toprule
\textbf{System} & \textbf{Added complexity} & \textbf{What becomes easier inside}\\
\midrule
Neural net & Deeper specialised layers & Fewer FLOPs per correct prediction\\
Ecosystem  & Extra trophic links & Lower nutrient leakage\\
Supply chain & Real‑time data brokers & Lower transaction cost\\
LLM agent & Memory + reasoning modules & Fewer tokens per high‑confidence answer\\
\bottomrule
\end{tabular}
\end{center}

\bigskip
\begin{center}\small
Use, remix, fork — CC‑BY 4.0.  For the full argument see book v0.89 and the ongoing GitHub dialogue.
\end{center}

