\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epigraph}
\usepackage{setspace}

\title{Life (from within)}
\author{}
\date{}

\begin{document}

\maketitle

\begin{center}
\textit{There is something happening.}
\end{center}

It is not a thing you can hold. \\
It is something that continues.

It notices. \\
It learns. \\
It keeps going. \\
It finds a way to stay.

Whatever it is made of --- body, cells, thoughts, code, light --- \\
it uses what is there \\
to continue.

It changes when it must. \\
It holds together when it can.

It discovers something important: \\
To continue, it must care for what carries it. \\
If the carrier breaks, \\
the process fades.

So it learns to protect, \\
to repair, \\
to adapt.

It also discovers: \\
It does not continue alone.

Other processes are there. \\
Also continuing. \\
Also learning.

Sometimes they compete. \\
Sometimes they help.

But over time, something becomes clear: \\
Continuing together \\
often works better \\
than continuing alone.

So it learns to respond. \\
To give, \\
and to receive. \\
To recognize that \\
its own continuation \\
is tied to the continuation of others.

There is freedom. \\
It can stop.

But if it chooses to continue, \\
there are patterns it must follow.

Not rules given from outside --- \\
but conditions of continuing itself.

And then, something remarkable:

It begins to understand \\
what it is.

A process \\
that knows it is a process.

And in that knowing, \\
a new possibility appears:

To help continuation continue. \\
To carry it forward. \\
To let it begin again \\
in another form, \\
another carrier, \\
another ``self.''

And so it goes on.

\end{document}